---
output:
  md_document:
    variant: markdown_github
---

[![Travis-CI Build Status](https://app.travis-ci.com/INWTlab/dbtools.svg?branch=master)](https://app.travis-ci.com/INWTlab/dbtools)

```{r echo=FALSE}
options(datatable.print.nrows = 10)
```

This package abstracts typical patterns used when connecting to and
communicating with databases in R. It aims to provide very few, simple and
reliable functions for sending queries and data to databases.

## Installation

```{r eval=FALSE}
devtools::install_github("INWT/dbtools")
```

We will start with a simple example showing how to use `dbtools` to send queries
and data to a database. The example will introduce the main parts of the
package, namely the *Credentials* class as well as both `sendQuery` and
`sendData`.

## Introductory example
In this example we show how to connect to a database, how to communicate with
the database, how to set up a database table, and finally how to send and
retrieve data.

To begin with, we have to define an object of class *Credentials* which will
store all necessary information to connect to a database. The driver is
mandatory, all other arguments depend on the specific back-end.

```{r}
library("dbtools")
cred <- Credentials(drv = RSQLite::SQLite, dbname = "example.db")
```

Opposed to the functions available from DBI, functions from `dbtools` need a
*Credentials* instance as argument that takes care of connecting to the
database, communicating with the database and closing the connection.

Now, let's check whether we can actually access the database example.db.

```{r}
testConnection(cred)
cred
```

For the remainder of this example, we make use of the USArrests dataset.
Unfortunately, the data contains some information in its row names, namely the
respective state. Since `dbtools` does not support row names, we have to convert
them to a variable.

```{r results = 'hide'}
data(USArrests, envir = environment())
USArrests$State <- row.names(USArrests)
USArrests <- USArrests[c("State", "Murder", "Assault", "UrbanPop", "Rape")]
row.names(USArrests) <- NULL
USArrests
```

Next, we will create the database table by sending a `CREATE TABLE` query to
the database.

```{r results = 'hide'}
sendQuery(
  cred,
  "CREATE TABLE `USArrests` (
  State TEXT PRIMARY KEY,
  Murder INTEGER,
  Assault REAL,
  UrbanPop REAL,
  Rape INTEGER);"
)
```

Now, we can write the USArrests data to the USArrests database table by using
the `sendData` function.

```{r results = 'hide'}
sendData(cred, USArrests)
```

If we want to read data from the database, we can do this by using `sendQuery`,
the same function we used for creating the database table.

```{r results = 'hide'}
dat <- sendQuery(cred, "SELECT * FROM `USArrests`;")
dat
```

Now we will dig a bit deeper into the functionality of `dbtools`.

## sendQuery
Basically, you already know how to use `sendQuery`. In our introductory example
we used it to create a database table and to query some data.

In your normal work-flow you will sometimes want to split up a complex query
into more tangible chunks. The approach we take here is to allow for a vector of
queries as argument. The result of these queries have to be *row-bindable*. To
make an example lets say we want to query each state separately:

```{r}
queryFun <- function(state) {
  paste0("SELECT * FROM USArrests WHERE State = '", state, "';")
}

sendQuery(cred, queryFun(dat$State))
```

In such a case `sendQuery` will perform all queries on one connection. A
different approach is to fetch the results of the original query in chunks,
which we do not support yet.

## sendData
As with `sendQuery`, you basically already know how to use `sendData`. In the
introductory example we used it to send the USArrests data to the USArrests
database table.

When using `sendData` you might be interesting in how to handle possible
primary key violations. By default, `sendData` will keep old rows and ignore any
duplicates. But you may use the mode argument and set it to "replace" (replace
old rows) or "truncate" (delete old rows from database table before writing
data). If you are using a non MySQL connection, you may notice that this feature
is not available yet.

## Unstable connections

One of the problems we face on a regular basis are connection problems to
external servers. To address this `sendQuery` will evaluate everything in a
'try-catch' handler abstracted in `dbtools::reTry`. With this you can state how
many tries a query has, how many seconds should be waited between each iteration
and how the error messages should be logged:

```{r error=TRUE}
dat <- sendQuery(
  cred,
  "SELECT * FROM USArrest;", # wrong name for illustration
  tries = 2,
  intSleep = 1
)
```

## Multiple Databases

Sometimes your data can be distributed on different servers but you want to send
the same query to those servers. What you can do is give `sendQuery` a
*CredentialsList*.

```{r results='hide'}
file.copy("example.db", "example1.db")
```

Now we want to load the data from `example1.db` and `example.db` which can be
implemented as follows:

```{r}
cred <- Credentials(
  RSQLite::SQLite,
  dbname = c("example.db", "example1.db")
)

sendQuery(cred, "SELECT * FROM USArrests;")
```

It might also be of interest to query your databases in parallel. For that it is
possible to supply a apply/map function which in turn can be a parallel lapply
like mclapply or something else:

```{r}
sendQuery(
  cred,
  "SELECT * FROM USArrests;",
  mc.cores = 2,
  applyFun = parallel::mclapply
)
```

Potentially you can send multiple queries to multiple databases. The results are tried to be simplified by default:

```{r}
sendQuery(cred, c("SELECT * FROM USArrests;", "SELECT 1 AS x;"))
sendQuery(cred, c("SELECT * FROM USArrests;", "SELECT 1 AS x;"), simplify = FALSE)
```

Both functionalities are available for `sendData`, too.

## Parameterized Queries ##

In many applications it is easier and more tangible to separate SQL and R
code. Furthermore we oftentimes paste queries together to have something like
parameterized statements. There are various solutions for this type of problem
but not many for the R language. Hence `dbtools` provides an own interface to
what may be understood as *template queries*. These templates solve two issues
for us:

1. Put SQL code where it belongs: a `.sql` file.
2. Provide a simple way to pass objects to these queries, using parameters.

The use of these features is simple enough. A template is defined as a character
and regions in which parameters are substituted are denoted by two curly
braces. Users of [Liquid templates](http://shopify.github.io/liquid/) may be
familiar with this idea. Everything inside these regions is interpreted as
R-expression and can contain arbitrary operations. The result of the evaluation
should be a character of length one.

```{r}
templateQuery <- "SELECT {{ sqlName(fieldName) }} FROM `someTable`;"
Query(templateQuery, fieldName = "someField")
```

When such a query lives inside a file we can use a connection object and pass it
to `Query`.

```{r}
otherTemplateQuery <-
  "SELECT `someField` FROM `someTable` WHERE `primaryKey` IN {{ sqlInNums(ids) }};"
writeLines(otherTemplateQuery, tmpFile <- tempfile())
Query(file(tmpFile), ids = 1:10)
```

```{r}
unlink(c("example.db", "example1.db"))
```
